A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-28 text/manuscript.Rmd
cut all text to conform to Cell's 45k character limit

To study cancer pathways we obtained a set of 7,607 solid tumor samples from The
Cancer Genome Atlas (TCGA) through the [Genomic Data Commons
(GDC)](gdc.cancer.gov) portal. TCGA data were most appropriate for our study
given the relative completeness of the patient metadata, particularly for
survival and staging. We chose to analyze somatic mutations in exome sequencing
data because the affected target gene is known unambiguously. Therefore, we
selected all missense, nonsense, frameshift, stop-loss, untranslated region, and
splicing mutations. In order to minimize bias from well-studied diseases and
processes, we selected 377 [Reactome pathways](https://reactome.org/) (see
supplemental table ST1) of interest corresponding to basic cellular processes
and biochemical pathways, excluding gene sets that correspond to miscellaneous
categories (_e.g._ “transcription factors”) or disease associations (_e.g._
“mutated in colon cancer”) and filtered our gene list on membership in these
pathways (total of 8,940 genes; see methods for details).

To avoid bias toward larger pathways (_i.e._ pathways with more member genes),
we counted pathways as disrupted if one or more member genes were mutated. We do
not attempt to calculate enrichment for mutations within a pathway. Data
binarized by pathway are likely to be noisy for several reasons. First, point
mutations can be deleterious (attenuating, hypomorphic or antimorphic) or
activating (neomorphic or hypermorphic) in genes, and these can in turn act
functionally as either oncogenes or tumor suppressors. For this study we assume
a significant fraction of these mutations are generically disruptive to normal
pathway activity since it is impossible to know the tumor promoting effects of
all mutations, including rarely studied genes. Second, we know that
low-expressed genes and non-expressed genes accumulate mutations at a higher
rate due to transcription coupled repair
[@kandothMutationalLandscapeSignificance2013;
@kimPancancerAnalysisSomatic2018;
@Pervasive_lesion_segregation; @expression-aware-annotation]. To address this
issue, we identified genes with low expression in each type of cancer and
eliminated them for that cancer type only (see methods for details). Highly
expressed genes could also have high mutation rates owing to transcription
induced mutagenesis [@parkGenomicEvidenceElevated2012]. We felt that such
mechanisms result in cell-type-specific biases that might be biologically
meaningful for predisposition to different classes of cancer however, and
therefore chose not to exclude these genes from our analysis. After selecting
our pathways and genes, we then compiled a matrix of the pathways, assigning a
Boolean value of 1 to each pathway with one or more genes mutated and 0 for all
others (**figure 1**).

@@

**Figure 1: Molecular pathway profiles of tumor samples with one or more
mutations.** Each of 377 selected Reactome pathways (rows) is classified as
disrupted if one or more genes is mutated in the tumor sample (columns) where
red and grey represents the pathway is mutated and not, respectively. Tumors are
grouped by tissue of origin using standardized abbreviations from the TCGA
project.

@@

Having defined tumors in terms of their pathway disruption profile, we sought to
understand whether different cancer types segregated into one or more
predominant classes. To our surprise, most cancer types were not heavily biased
in one class, and all well-represented cancer types had tumors in every class
(see **figure 3A** and full tumor profiles in **supplementary figure S1**, see
also interactive media from **supplementary file 1**), suggesting that,
in principle at least, our pathway-disruptions identify clusters of molecular
pathology largely independent of tissue-of-origin. As an example of one type of
cancer that does have a biased pathway profile, pancreatic adenocarcinoma (PAAD)
was predominantly found in class 8 (**figure 3A** and **supplementary figure
S1** and **supplementary file 1**). But even PAAD comprises tumors from the nine remaining classes, meaning
that patients suffering these tumors have potentially different underlying
molecular pathologies.

@@

Many cancers have molecular or histological subtypes defined based on gene
expression or other -omics profiles or pathology lab results. These subtypes
often have different standards of care owing to different overall drug
sensitivity (or other factors). If the histological subtypes represent true
molecular phenotypes, one predicts that histological subtypes should segregate
with our pathway-based clusters, therefore providing support for the clusters as
proxies for molecular pathology sub-typing. To our surprise, we find a similar
result to the previous analysis of cancer types projected onto the UMAP of
pathway disruptions. To illustrate this, we projected annotations for each of
the breast cancer subtypes, composed of Triple-negative/Basal-like, Her2
positive, normal-like, and luminal A and B subtypes onto the UMAP. These are
among the most heavily studied molecular subtypes in cancer, which each have
different prognoses and standards of care. We did not observe any exclusive
segregation by pathway for these subtype annotations (**Figure 3B**). We also 
projected histological subtype data for the remaining cancers (see 
**supplementary figure S2** for the full set of projections and see interactive
media from [**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)); we find that the subtypes, though often 
biased towards one or more classes, are almost never exclusive. We interpret 
these data in aggregate to mean that our pathway disruption classes do not 
correspond to previously identified molecular subtypes within the parent cancer type.

@@

There are several well-known familial cancer-causing mutations that have been
interrogated extensively for differences in basic biology, survival and
treatment outcomes. However, the functions of these genes are related to risk
factors such as genome stability generally, proof-reading and DNA damage repair,
and telomere length. _BRCA1/2_ genes for example are key for DNA double-stranded
break repair [@moynihan-1999-brca1;@davies-2001-brca2] and germline mutations in
these genes confer elevated risk for breast, prostate and ovarian cancers. The
mechanism of risk is thought to involve loss of heterozygosity resulting in loss
of the wildtype, functional allele [@brca-loh], so we projected the somatic
mutations for _BRCA1_ and _BRCA2_ genes onto the UMAP, but did not observe
segregation of these mutations into specific clusters (**figure 3C** and see 
interactive media from [**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). We made similar projections 
for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH1_, _MLH3_, _PMS1_ 
and _PMS2_; _BRIP1_, _RAD51_, _CHEK2_ and _APC_. None of these genes except 
for _APC_ exhibited any remarkable specificity with respect to cluster assignment 
(**supplementary figure S3**). To look at other risk factors
such as maintenance of DNA methylation levels and telomere length, we projected
somatic mutations of the _TET2_ and _TET3_ genes, plus _TERT_, _TEP1_, and
_DKC1_, and observed similar lack of segregation by cluster (**supplementary
figure S3**).

@@

Tumor staging is based on physico-pathological criteria, including tumor
diameter, which can vary greatly in importance between different tissues. Stage
is used clinically as a proxy for advancement toward a more deadly state and
metastasis. Given these assumptions, it is possible that more advanced tumors
have common pathway disruption profiles. The UMAP, which features a series of
lobe-like structures on a common backbone of tumor samples (**figure 2**) could
in principle reflect progression through a series of stages. The backbone starts
with a cluster of tumors (class 1) that has the fewest point mutations and
culminates in a cluster (class 10) which has nearly every pathway disrupted
(**figure 4**). However, outside of class 10 we don't observe an obvious trend
in the overall mutation burden across the backbone of the UMAP. Nonetheless, to
test the hypothesis that the molecular-pathway disruption clusters represent
advancement through stages, we projected staging data onto the UMAP. Similar to
tissue of origin and other categories of tumor, we did not observe any bias
among the stages to specific pathway disruption clusters (**figure 3D**),
suggesting that stage is not a contributing factor to cluster identity.

@@

**Figure 4: Somatic mutation frequencies for tumors in each class** 
Vertical axis shows log mutation count, horizontal axis is cluster identity.
Each dot represents an individual tumor sample, ranked lowest to highest by
mutation count. The median mutation count in each cluster is indicated by the
horizontal line.

@@

Finally, as a measure of tumor advancement, metastasis is the condition in which
certain phenotypic criteria are met. These phenotypes include loss of
differentiation, cell-cell contacts, epithelial to mesenchyme transition, immune
system evasion and tissue invasiveness [@hallmarks_II]. To determine whether any
of our clusters correspond to an especially advanced stage of cancer across
tissue types, we projected the metastasis data onto our UMAP, and surprisingly
we observed an even distribution of metastases across classes (**figure 3E**).
This final observation suggests that our pathway-disruption classification is
dependent on particular combinations of gene mutations affecting different
pathways that can each give rise to advanced stages of disease and metastasis,
regardless and independent of overall mutational burden.

@@

In order to identify pathway disruption enrichment across all cancers
(pan-cancer), we created a list of pathway disruptions with percent mutated
samples and top genes (**supplementary table ST1**). As expected,
these analyses reveal the broad importance of many well known pathways that are
disrupted in cancer, including "PIP3 activates Akt signaling" (77% of samples),
"MAP1K/MAP3K signaling" (70% of samples), "Mitotic G2-G2/M phases" (67% of
samples), "Cellular senescence" (64% of samples), "G2/M Checkpoints" (62% of
samples), _etc._

To discover what pathways are most important for clustering, we calculated
percent enrichment for each pathway _within cluster_ relative to all other
clusters combined (see methods) and ranked pathways for each cluster from
highest to lowest enrichment. We visualized the enrichment as a heatmap
(**figure 5A**). Using this approach, we identified twenty-five pathways highly
enriched (enrichment score ≥ 0.3, 95% confidence; see methods) in cluster 2,
nine pathways enriched in cluster 3, twenty-four pathways in cluster 4, ten
pathways in cluster 5, ten pathways in cluster 6, twenty-four pathways in
cluster 7, twenty-four pathways in cluster 8, ten pathways in cluster 9, and no
pathways for clusters 1 and 10 (**supplementary table ST2**). Clusters 7, 8, and
9 had several pathways in common. To explore the specific pathways marking each
cluster, we projected disruptions for each of the 377 pathways onto the UMAP
(**supplementary figure S4** and see interactive media from [**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). 
Clusters 3 and 5 were distinguished by metabolic pathways including RNA and 
protein biosynthesis (**supplementary figure S4**). Similarly, cluster 4 was 
distinguished by mutations affecting regulation of DNA
and histone methylation ("DNA methylation", "PRC2 methylates histones and DNA",
and "Nucleosome assembly"). Clusters 7-9 have in common mutations in
extracellular, intracellular, and immune-related signaling pathways (see
**figure 5B** and **supplementary figure S4**). Cluster 2 had the highest
pathway enrichment levels of the three, having mutations in hedgehog signaling,
"β-catenin degradation", "cellular response to hypoxia", "regulation of cell
cycle" and "apoptosis" among others.

@@

Prior efforts to extract common signatures from pan-cancer datasets met with
difficulty in distinguishing tumor samples from their tissue-specific -omics
data signatures. Given our pathway-disruption based clustering, this raises
the question, are tumor phenotypes driven entirely by common driver genes, or by
“silent” tissue-specific effectors (_i.e._ too few samples to detect above
statistical significance thresholds), or a combination of both? To answer this
question, we compared top pathway genes for each cluster relative to the TCGA
background to find differentially mutated genes. We ranked odds ratios and
selected the top ten enriched and depleted genes (pvalue < 0.01) for each
cluster (**Figure 6**; odds ratios plot). Clusters 7 and 8, which shared
multiple enrichment in signaling pathways, are largely driven by mutations in
_PI3K_ and its orthologs and _Ras_ genes, respectively (compare _PIK3CA_ and
_KRAS_ panels of **Supplementary figure S5** and see interactive media from 
[**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). Interestingly, cluster 9, which
also shared multiple enrichment in signaling pathways with clusters 7 and 8, is
enriched for both _PIK3CA_ and _KRAS_. Clusters 3 and 5, defined by enrichment
in metabolic pathways, had mutations in ribosomal proteins and nuclear pore
complex, respectively. Cluster 4 had mutations in genes responsible for
nucleosome structure. Cluster 2 had mutations in proteasomal subunit genes
involved in protein degradation. We also observed that genes that were enriched
for one cluster are depleted from others (_i.e._ is enriched in cluster 6,
but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but depleted in
clusters 3 and 8). Next, we investigated the proportion of samples per cancer
type for the significant genes within a cluster (**Figure 6**; heatmap).
Surprisingly, cluster-specific tumors were not predominated by one or more
highly mutated genes across all cancers. Instead, when observing the mutation
rate for these genes within samples that belong to a cluster, the mutation rate
is heterogeneous across tumors by tissue origin (_e.g._ In cluster 4, CESC was
enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was enriched for
_HIST1H2AC_). Even among the top most mutationally enriched genes within
clusters there is no global pattern, indicating that our clusters are not driven
by individual genes, but rather networks as a whole. Taken together, our data
identifies a framework of cancer type-specific mutations associated with
specific clusters.

@@

**Figure 6: Gene level analysis reveals tissue-specific class signatures.** 
Odds ratio plot; Column uses a logarithmic axis to represent odds ratio with a
95% confidence interval. Row represents the significant genes from each cluster.
Each cluster was compared against the background (all other clusters) to find
differentially mutated genes. Significant genes (pvalue < 0.01) were selected
and limited to the top ten results for each clusters. Heatmap; Column represents
the cancer type. Row represents the significant genes as described above. The
heatmap shows the proportion of samples per cluster and cancer type that are
mutated for each gene. Depleted significant genes (left). Enriched significant
genes (right).

@@

Following the same logic we used to investigate cluster-specific enrichment of
pathways we compared metastatic _vs._ non-metastatic tumors, as we find that
metastatic tumors are distributed across all ten clusters (_e.g._ **figure
3E**). Using all non-metastatic tumors as background, we found very low levels
of enrichment (< 10%) in a handful of pathways. We reasoned that the individual
clusters might be too different to detect global metastasis enrichment signal
given the small sample size (n = 215 metastatic tumor samples).

Therefore, we calculated cluster-specific enrichment in metastatic tumors and
found a total of 31 enriched pathways (significant with enrichment score ≥ 0.3)
across all clusters (**table 1**). A number of enrichments were found in
multiple clusters represented by pathways that were already shown to be enriched
in a neighboring cluster. For example, “Signaling by _PTK6_” is enriched in
non-metastatic samples of cluster 8 (see **supplementary figure S4**) but not in
7 and 9. This pathway is enriched in metastatic tumors of clusters 7 and 9 ($p <
10^{-3}$, **table 1**). This is also true of “Erythropoietin activates _RAS_”,
which is enriched in non-metastatic tumors of cluster 8 (**supplementary figure
S4**) and also in metastatic tumors of clusters 7 and 9. Cluster 4 metastatic
samples were enriched for “Fc epsilon receptor (FCERI) signaling”, a key
neutrophil pathway, which is also specific to clusters 2, 7 and 8 non-metastatic
tumors. Thus, metastatic-enriched pathways from one cluster are often enriched in 
non-metastatic tumors of other clusters.

@@

All code for producing the analyses and figures herein are included in this
fully reproducible manuscript in R markdown format. R markdown files and all
other models are available from [our
repositories](https://github.com/dennishazelett/TTmanu) on the distributed
version control site, GitHub.

### Selection of pathways

To understand the molecular mechanism of cancer at a pathway level, we used
Reactome (https://reactome.org/), a knowledge-based pathway database. The
mapping files of ENSEMBL genes to pathways, pathway hierarchy relationships, and
complexes to top pathways were downloaded from
https://reactome.org/download-data. Using these data, we imposed pathway
criteria to define basic cellular processes and biochemical pathways: (1)
human-derived pathways (\it{“HSA”}) (2) limited to grandchild node for each
parent pathway (_e.g._ ‘Beta-catenin independent WNT signaling’ in ‘Signal
Transduction’) (3) exclusion of pathways in the parent pathway: "Disease",
"Muscle contraction", and "Reproduction" or pathway names that include any of
the following keywords: "disease", "defect", "cancer", "oncogenic", "disorder",
"mutant", "loss", "infection", "bacteria", or "listeria". While some of the
excluded pathways have been shown to play an important role in cancer, they are
highly specialized (_e.g._ "PI3K/AKT Signaling in Cancer"). Additionally, for
most of the excluded pathways, a neutral version pathway of the pathway exists
(_e.g._ "PIP3 activates Akt signaling"). Finally, we mapped the 18,577
Ensembl IDs from the TCGA dataset to the highly selected Reactome pathways. This
operation produced a lookup table that consisted of 377 pathways mapped to 8,940
genes.

### Filtering genes

We filtered likely erroneous mutations due to transcription coupled repair. Our
approach was to determine the status of each gene (_i.e._ expressed or not
expressed) in each tissue type in order to exclude gene mutations low expressed
enes. To do this, we obtained the TCGA RNA sequencing data adjusted for batch
effect dataset (https://pancanatlas.xenahubs.net). Using the data, we removed
the genes and tumor samples that were not included in our analysis, grouped the
tumor samples by tissue type and computed the mean expression value for each
gene. A minimum threshold of 10 transcripts per million was set for expressed
genes based on an inflection point observed when plotting the mean expression
values of genes ranked by expression in each tissue type. Genes that did not
meet this threshold were considered not expressed. This operation produced a
lookup table for gene expression status in each tumor sample for 18,127 genes.

### Class and stage-specific Enrichment Calculations

We calculated the enrichment of pathways in one set of tumors as the relative
fraction of tumors (with estimated uncertainty) with a mutated gene in the
pathway to all tumors, inclusive of the category of interest. To do this
operation, we use the beta distribution to permute a posterior distribution on
the fraction of tumors with a pathway mutated for each category based on the
observed set, and compared this to the posterior obtained from the full set of
tumors (all tumors) as the distribution of differences between all permuted
samples. We considered a pathway enriched if the 95% range of credible
differences thus obtained excludes 0 and the mean credible difference was
greater than or equal to 30% enrichment, which excludes a large number of small
differences that are not likely to be biologically relevant.

### Models of survival based on population-specific longevity data

We used Bayesian inference to explore the relative impact of cancer
diagnosis on survival at the time of diagnosis. Our model assumes that
at birth each person has a small initial probability per unit time to
die, and that this probability grows exponentially in time at some
baseline rate that matches the observed CDC data. The rate of growth
in risk is assumed to jump to a higher level at diagnosis, which we
take as a proxy for onset. This represents an accelerated aging type
model. To fit the model, we constrain the baseline risks using the
available CDC life tables
(ftp://ftp.cdc.gov/pub/Health_Statistics/NCHS/Publications/NVSR/68_07/). We
also constrain the cancer specific multiplier using the longevity data
within the cancer tumor dataset. Our model estimates the effect of
cancer type and class-specific cancer effects independently, resulting
in a cancer and class specific estimate of the aging rate multiplier.
A detailed description of the CDC longevity model and likelihood is
provided in the **Supplementary methods**.

>>>>

To study cancer pathways we obtained a set of 7,607 solid tumor samples from The
Cancer Genome Atlas (TCGA) through the [Genomic Data Commons
(GDC)](gdc.cancer.gov) portal. TCGA data were most appropriate for our study
given the relative completeness of the patient metadata, particularly for
survival and staging. We chose to analyze somatic mutations in exome sequencing
because the affected target gene is known unambiguously. Therefore, we
selected all missense, nonsense, frameshift, stop-loss, untranslated region, and
splicing mutations. In order to minimize bias from well-studied diseases and
processes, we selected 377 [Reactome pathways](https://reactome.org/) (see
supplemental table ST1) of interest corresponding to basic cellular processes
and biochemical pathways, excluding gene sets that correspond to miscellaneous
categories (_e.g._ “transcription factors”) or disease associations (_e.g._
“mutated in colon cancer”) and filtered our gene list on membership in these
pathways (total of 8,940 genes; see methods for details).

To avoid bias toward larger pathways with more genes, we counted them as
disrupted if one or more member genes were mutated. We do not attempt to
calculate enrichment for mutations within a pathway. Data binarized by pathway
are likely to be noisy for several reasons. First, point mutations can be
deleterious (attenuating, hypomorphic or antimorphic) or activating (neomorphic
or hypermorphic) in genes, and these can in turn act functionally as either
oncogenes or tumor suppressors. For this study we assume a significant fraction
of these mutations are generically disruptive to normal pathway activity since
it is impossible to know the tumor promoting effects of all mutations, including
rarely studied genes. Second, we know that low- and non-expressed genes
accumulate mutations at a higher rate due to transcription coupled repair
[@kandothMutationalLandscapeSignificance2013;
@kimPancancerAnalysisSomatic2018;
@Pervasive_lesion_segregation; @expression-aware-annotation]. To address this
issue, we identified low expressed genes in each type of cancer and eliminated
them for that cancer type only (see methods for details). Highly expressed genes
could also have high mutation rates owing to transcription induced mutagenesis
[@parkGenomicEvidenceElevated2012]. This phenomenon would likely result in
cell-type-specific biases that might be biologically meaningful for
predisposition to different classes of cancer however, and therefore chose not
to exclude these genes from our analysis. After selecting our pathways and
genes, we then compiled a matrix of the pathways, assigning a Boolean value of 1
to each pathway with one or more genes mutated and 0 for all others (**figure
1**).

@@

**Figure 1: Molecular pathway profiles of tumor samples with one or more
mutations.** Each of 377 selected Reactome pathways (rows) is classified as
disrupted if one or more genes is mutated in the tumor sample (columns). Red
cells denote pathway disruption. Tumors types reflect standardized abbreviations
from the TCGA project.

@@

Having defined tumors in terms of their pathway disruption profile, we sought to
understand whether different cancer types segregate into one or more predominant
clusters. To our surprise, most cancer types were not heavily biased in one
cluster, and all well-represented cancer types had tumors in every cluster (see
**figure 3A** and full tumor profiles in **supplementary figure S1**, see also
interactive media from **supplementary file 1**), suggesting that our
pathway-disruptions identify clusters of molecular pathology largely independent
of tissue-of-origin. As an example of one type of cancer that does have a biased
pathway profile, pancreatic adenocarcinoma (PAAD) was predominantly found in
cluster 8 (**figure 3A** and **supplementary figure S1** and **supplementary
file 1**). But even PAAD comprises tumors from the nine remaining classes,
meaning that patients with these tumors have potentially different underlying
molecular pathologies.

@@

Many cancers have molecular or histological subtypes defined based on gene
expression, pathology or other -omics profiles. These subtypes often have
different standards of care owing to different drug sensitivity (or other
factors). If the histological subtypes represent true molecular phenotypes, one
predicts that they should segregate with our pathway-based clusters, therefore
providing support for the clusters as proxies for molecular pathology
sub-typing. To our surprise, we find a similar result to the previous analysis
of cancer types projected onto the UMAP of pathway disruptions. To illustrate
this, we projected annotations for each of the breast cancer subtypes, composed
of Triple-negative/Basal-like, Her2 positive, normal-like, and luminal A and B
subtypes onto the UMAP. These are among the most heavily studied molecular
subtypes in cancer, each with different prognoses and standards of care. We did
not observe any exclusive segregation by pathway for these subtypes (**Figure
3B**). We also projected histological subtype data for the remaining cancers
(see **supplementary figure S2** and online [**supplemental file
1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)); we find that the
subtypes, though often biased to one or more clusters, are almost never
exclusive. We interpret these data in aggregate to mean that the clusters do not
correspond to previously identified molecular subtypes within the parent cancer
types.

@@

There are several well-known familial cancer-causing mutations that have been
studied for differences in basic biology, survival and treatment outcomes. The
functions of these genes are related to risk factors such as genome stability,
proof-reading & DNA damage repair, and telomere length. _BRCA1/2_ genes for
example are key for DNA double-stranded break repair
[@moynihan-1999-brca1;@davies-2001-brca2] and germline mutations in these genes
confer elevated risk for breast, prostate and ovarian cancers. The mechanism of
risk is thought to involve loss of heterozygosity resulting in loss of the
wildtype, functional allele [@brca-loh], so we projected the somatic mutations
for _BRCA1_ and _BRCA2_ genes onto the UMAP. We did not observe segregation of
these mutations into specific clusters (**figure 3C** and see interactive media
from [**supplemental file
1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). We made similar
projections for the mismatch repair (MMR) genes _MSH2_, _MSH6_, _MLH1_, _MLH3_,
_PMS1_ and _PMS2_; _BRIP1_, _RAD51_, _CHEK2_ and _APC_. None of these genes
except for _APC_ exhibited any remarkable specificity with respect to cluster
assignment (**supplementary figure S3**). To look at other risk factors such as
maintenance of DNA methylation levels and telomere length, we projected somatic
mutations of the _TET2_ and _TET3_ genes, plus _TERT_, _TEP1_, and _DKC1_, and
observed similar lack of segregation by cluster (**supplementary figure S3**).

@@

Tumor staging is based on physico-pathological criteria, including tumor
diameter, which can vary greatly in importance between different tissues. Stage
is used clinically as a proxy for advancement toward a more deadly state and
metastasis. Given these criteria, it is possible that more advanced tumors have
common pathway disruption profiles. The UMAP, which features a series of
lobe-like structures on a common backbone of tumor samples (**figure 2**) could
in principle reflect progression through a series of stages. The backbone starts
with a cluster of tumors (class 1) that has the fewest point mutations and
culminates in a cluster (class 10) which has nearly every pathway disrupted
(**figure 4**). However, outside of class 10 we don't observe a trend in the
overall mutation burden across the backbone of the UMAP. Nonetheless, to test
the hypothesis that the molecular-pathway disruption clusters represent
advancement through stages, we projected staging data onto the UMAP. Similar to
tissue of origin and other categories of tumor, we did not observe any bias
among the stages to specific clusters (**figure 3D**), suggesting that stage is
not a contributing factor to cluster identity.

@@

**Figure 4: Somatic mutation frequencies for each cluster** 
Vertical axis shows log mutation count, horizontal axis is cluster identity.
Each dot represents an individual tumor sample, ranked lowest to highest by
mutation count. The median mutation count in each cluster is indicated by the
horizontal line.

@@

Finally, as a measure of tumor advancement, metastasis is the condition in which
certain phenotypic criteria are met: loss of
differentiation, cell-cell contacts, epithelial to mesenchyme transition, immune
system evasion and tissue invasiveness [@hallmarks_II]. To determine whether any
of our clusters correspond to an especially advanced stage of cancer across
tissue types, we projected the metastases onto our UMAP, and surprisingly
we observed even distribution of the samples across clusters (**figure 3E**).
This final observation suggests that our pathway-disruption clustering is
dependent on particular combinations of gene mutations affecting different
pathways that can each give rise to advanced stages of disease and metastasis,
regardless and independent of overall mutational burden.

@@

In order to identify pathway enrichment across all cancers (pan-cancer), we
created a list of pathway disruptions with percent mutated samples and top genes
(**supplementary table ST1**). As expected, these analyses reveal the broad
importance of many well known pathways that are disrupted in cancer, including
"PIP3 activates Akt signaling" (77% of samples), "MAP1K/MAP3K signaling" (70% of
samples), "Mitotic G2-G2/M phases" (67% of samples), "Cellular senescence" (64%
of samples), "G2/M Checkpoints" (62% of samples), _etc._

To discover what pathways are most important for clustering, we calculated
percent enrichment _within cluster_ relative to all other clusters (see methods)
and ranked pathways from highest to lowest enrichment. We visualized the
enrichment as a heatmap (**figure 5A**). Using this approach, we identified
about fourteen pathways per cluster (enrichment score ≥ 0.3, 95% confidence; see
methods) (**supplementary table ST2**). Clusters 7, 8, and 9 had several
pathways in common. To explore the specific pathways marking each cluster, we
projected disruptions for each of the 377 pathways onto the UMAP
(**supplementary figure S4** and see interactive media from [**supplemental file
1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). Clusters 3 and 5
were distinguished by metabolic pathways including RNA and protein biosynthesis
(**supplementary figure S4**). Similarly, cluster 4 was distinguished by
mutations affecting regulation of DNA and histone methylation ("DNA
methylation", "PRC2 methylates histones and DNA", and "Nucleosome assembly").
Clusters 7-9 have in common mutations in extracellular, intracellular, and
immune-related signaling pathways (see **figure 5B** and **supplementary figure
S4**). Cluster 2 had the highest pathway enrichment levels of the three, having
mutations in hedgehog signaling, "β-catenin degradation", "cellular response to
hypoxia", "regulation of cell cycle" and "apoptosis" among others.

@@

Prior efforts to extract signatures from pan-cancer datasets met with difficulty
in distinguishing tumor samples from their tissue-specific -omics data
signatures. Given our pathway-disruption based clustering, this raises the
question, are tumor phenotypes driven entirely by common driver genes, or by
“silent” tissue-specific effectors (_i.e._ too few samples to detect above
statistical significance thresholds), or a combination of both? To answer this
question, we compared top pathway genes for each cluster relative to the TCGA
background to find differentially mutated genes. We ranked odds ratios and
selected the top ten enriched and depleted genes (pvalue < 0.01) for each
cluster (**Figure 6**; odds ratios plot). Clusters 7 and 8, which shared
multiple enrichment in signaling pathways, are largely driven by mutations in
_PI3K_ and its orthologs and _Ras_ genes, respectively (compare _PIK3CA_ and
_KRAS_ panels of **Supplementary figure S5** and see interactive media from
[**supplemental file
1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). Interestingly,
cluster 9, which also shared multiple enrichment in signaling pathways with
clusters 7 and 8, is enriched for both _PIK3CA_ and _KRAS_. Clusters 3 and 5,
defined by enrichment in metabolic pathways, had mutations in ribosomal proteins
and nuclear pore complex, respectively. Cluster 4 had mutations in genes
responsible for nucleosome structure. Cluster 2 had mutations in proteasomal
subunit genes involved in protein degradation. We also observed that genes that
were enriched for one cluster are depleted from others (_i.e._ is enriched in
cluster 6, but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but
depleted in clusters 3 and 8). Next, we investigated the proportion of samples
per cancer type for the significant genes within a cluster (**Figure 6**;
heatmap). Surprisingly, cluster-specific tumors were not predominated by one or
more highly mutated genes across all cancers. Instead, when observing the
mutation rate for these genes within samples that belong to a cluster, the
mutation rate is heterogeneous across tumors by tissue origin (_e.g._ In cluster
4, CESC was enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was
enriched for _HIST1H2AC_). Even among the top enriched genes within clusters
there is no global pattern, indicating that our clusters are not driven by
individual genes, but rather networks as a whole. Taken together, our data
identify a framework of cancer type-specific mutations associated with specific
clusters.

@@

**Figure 6: Gene level analysis reveals tissue-specific class signatures.** 
Odds ratio plot; Column uses a logarithmic axis to represent odds ratio with a
95% confidence interval. Rows: significant genes from each cluster. Each cluster
was compared against the background (all other clusters) to find differentially
mutated genes. Significant genes (pvalue < 0.01) were selected and limited to
the top ten results for each cluster. Heatmap; Columns: cancer type. Rows:
significant genes. The heatmap shows proportion of samples per cluster and
cancer type mutated for each gene. Depleted significant genes (left). Enriched
significant genes (right).

@@

Since metastatic tumors are distributed across all ten clusters, we first
compared them with non-metastatic tumors following the same logic we used to
investigate cluster specific enrichment. Using all non-metastatic tumors as
background, we found very low levels of enrichment (< 10%) in a handful of
pathways. We reasoned that the individual clusters might be too different to
detect global metastasis enrichment signal given the small sample size (n = 215
metastatic tumor samples).

Therefore, we calculated cluster-specific enrichment in metastatic tumors and
found a total of 31 enriched pathways (significant with enrichment score ≥ 0.3)
across all clusters (**table 1**). A number of enrichments represented pathways
that were already shown to be enriched in non-metastatic samples of other
clusters. For example, “Signaling by _PTK6_” is enriched in non-metastatic
samples of cluster 8 (see **supplementary figure S4**) but not in 7 and 9. This
pathway is enriched in metastatic tumors of clusters 7 and 9 ($p < 10^{-3}$,
**table 1**). This is also true of “Erythropoietin activates _RAS_”, which is
enriched in non-metastatic tumors of cluster 8 (**supplementary figure S4**) and
also in metastatic tumors of clusters 7 and 9. Cluster 4 metastatic samples were
enriched for “Fc epsilon receptor (FCERI) signaling”, a key neutrophil pathway,
which is also specific to clusters 2, 7 and 8 non-metastatic tumors. Thus,
metastatic-enriched pathways from one cluster are often enriched in
non-metastatic tumors of other clusters.

@@

All code and data for producing the analyses and figures herein are included in
this fully reproducible manuscript in R markdown format. R markdown files and
all other models are available from [our
repositories](https://github.com/dennishazelett/TTmanu) on the distributed
version control site, GitHub. Extended descriptions of clustering, models,
formulae, pathway selection, and gene filtering are provided in supplementary
methods.
